---
title: "refactor"
author: "Lorenz Walthert and Jon Calder"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{refactor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Working with factors in R can be frustrating. As described well by [John Monut](http://www.win-vector.com/blog/2014/09/factors-are-not-first-class-citizens-in-r/), factors are not first-class citizens in R. Even most basic code does often behave in an unexpeced manner and without any warning. Let's start off with a simple example that illustrates some of the issues: Creating two factor vectors and combining them together:

```{r}
lower <- factor(letters[1:3], levels = letters[2:4])
upper <- factor(LETTERS[1:5], levels = LETTERS[1:4])
(combined <- c(lower, upper))
```


One can note the following issues:

* `factor()` does not yield an error when `NA`s are created silently, neither when levels does not match x nor when too few levels are specified. 
* When `c()` is used to combine the two, the levels of the two elements (which are just attributes to integer vectors) are not conveyed, R puts together the underlying integers.
* This is not only changing the data representation but also make it impossible to restore the information, i.e. 1 represents now the first level of both lower and upper, so it can either be `b` or `A`. 

## `cfactor`
`cfactor` has been defined to give warnings in cases when empty factors are created from strings or when existing strings are not preserved. Essentially, it is a wrapper for `factor()`, and the 'c' stands for enhanced control. It also has an improved level order detection based on numerical values within strings that is superior to the way `factor` assesses this order in certain cases.

### unmatched factors
```{r}
string <- c("a", "b", "c")
library(refactor)
# with refactor::cfactor warnings are issued to communicate fact that levels were referenced that do not appear in the data and vice versa
cfactor(string, levels = c("b", "c", "d"))
```

### detect levels
`cfactor` should detect sequences in strings if the argument `ordered` is set to `TRUE` order the levels. The default behavior of `factor` is
`factor(x, levels = sort(unique(as.character(x)))`, which would be fine if the string ordering was the same as ordering the numbers within the string, but in general that is often not the case.

\
**A simple example**
```{r}
money <- c("EUR 11 - EUR 20", "EUR 1 - EUR 10", "EUR 21 - EUR 22") # no issue, since all numbers are two digits
factor(money, ordered = T) # correctly detects level

money <- c("EUR 21 - EUR 22", "EUR 100 - 101", "EUR 1 - EUR 10", "EUR 11 - EUR 20")  # the general case
cfactor(money, ordered = T) # cfactor detects levels using regular expressions
```

\
**A more "dirty" example**
```{r}
dirty_money <- c("EUR 21 - EUR 22", "EUR 100 - 101", "EUR 11 bis EUR 20", "from 98 to 99")
cfactor(dirty_money, ordered = T)
```

\
**Other example strings**
```{r, echo = F, include = T}
c("1 to 4", "5 to 6") # properly separated
c("1 to 4", "4 to 6") # not properly separated
c("from 1,000 to 2,000", "from 2000 to 4,000") # comma separated and 'from' and 'to'
c("4.0 / 4.1", "4.2 / 4.3") # point and slash separator
c("one minute", "three minutes", "1 hour")
```


### boundary labels for integer bins
using `cut` gives boundaries like [50, 55) [55, 60), even for integers. This is undesired in the case of integer values, so it should give boundaries like "50 - 55", "56 - 60". Possibly with the spacing being optional.

```{r}
random <- sample(100)
cut.default(random, breaks = seq(0, 100, by = 10))[1:10] # the uggly outcome with cut.default

class(random) # ok, we define an S3 method for integer

library(refactor)
cut(random, breaks = seq(0, 100, by = 10))[1:10]
```

### cut method for ordered factors
This method should allow to summarize levels of factors into fewer categories given some break points
```{r}
library(refactor)
timeX <- c("very low", "rather low", "low", "medium", "upper medium", "high")
time <- factor(timeX, levels = unique(timeX), ordered = T)

cut(time, breaks = c("low", "medium", "high"))
```
